---
title: "scIGD on 20k PBMC"
author: "Ahmad Al Ajami"
date: "`r Sys.Date()`"
output:
  html_document:
    keep_md: false
    toc: true
    toc_float: true
    toc_depth: 5
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, include=TRUE, message=FALSE, cache = FALSE}
suppressPackageStartupMessages({
library(SingleCellAlleleExperiment)
library(tidyverse)
library(gridExtra)
library(DropletUtils)
library(BiocParallel)
library(scDblFinder)
library(Matrix)
library(AnnotationHub)
library(cowplot)
library(pheatmap)
library(batchelor)
})
```

# SingleCellAlleleExperiment

```{r scae, message=FALSE}
read_scae <- function(folder_path){
  counts <- read_allele_counts(folder_path,
                               lookup_file = read.csv(file.path(folder_path, "lookup_table_HLA.csv")),
                               sample_names = "pbmc_20k",
                               filter_mode = "yes",
                               log = TRUE,
                               gene_symbols = TRUE)
  return(counts)
}

scae <- read_scae(file.path("../data/scIGD/")) 
dim(scae)
```

# Pre-processing workflow

## Looking at distributions

Plotting the normalized expression of alleles:

```{r, allele_expression}
immune_features <- rownames(scae[rowData(scae)$NI_I == "I", ])

plotExpression(scae, immune_features) +
  theme_bw() +
  theme(axis.text = element_text(size = 5),
        axis.title = element_text(size = 8, face = "bold"), 
        axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
total_counts <- colSums(counts(scae))
  
ggplot(data.frame(Total_Counts = total_counts), aes(x = Total_Counts)) +
  geom_histogram(binwidth = 500, fill = "skyblue", color = "black") +
  labs(x = "total counts per cell",
       y = "number of cells") +
  theme_bw()
```

## Cell filtering

### Removing within-sample doublets

```{r doublets, message=FALSE, warning=FALSE}
set.seed(42L)

scae <- scDblFinder(scae, BPPARAM=MulticoreParam(4))
tbl <- table(scae$scDblFinder.class)
paste(names(tbl), tbl, sep = " ")
scae <- scae[, scae$scDblFinder.class == "singlet"]
```

### Removing low quality cells

```{r}
rownames(scae) <- sub("\\..*", "", rownames(scae))

# identifying the mitochondrial transcripts:
query(AnnotationHub(), "EnsDb.Hsapiens.")

# release 93 from cellranger
edb <- AnnotationHub()[["AH64446"]]
```

Some QC:

```{r, warning=FALSE}
qc_annotationhub <- function(scae){
  location <- mapIds(edb, keys=rownames(scae), keytype="GENEID", column="SEQNAME")
  is.mito <- which(location=="MT")
  df <- perCellQCMetrics(scae, subsets=list(Mito=is.mito))
  reasons <- quickPerCellQC(df, percent_subsets=c("subsets_Mito_percent"))
  tbl <- colSums(as.matrix(reasons))
  cat(unique(scae$state), "\n", paste(names(tbl), tbl, sep = " "), "\n")
  colData(scae) <- cbind(colData(scae), df)
  scae$discard <- reasons$discard
  return(scae)
}

scae <- qc_annotationhub(scae)
```

```{r, fig.width=10, fig.height=2}
gridExtra::grid.arrange(
  plotColData(scae, x="Sample", y="sum", colour_by="discard") +
    scale_y_log10() + ggtitle("Total count"),
  plotColData(scae, x="Sample", y="detected", colour_by="discard") +
    scale_y_log10() + ggtitle("Detected features"),
  plotColData(scae, x="Sample", y="subsets_Mito_percent", colour_by="discard") +
    ggtitle("Mito percent"),
  ncol=3)
```

```{r}
plotColData(scae, x="sum", y="subsets_Mito_percent", colour_by="discard") +
  theme(panel.border = element_rect(color = "grey"))
```

Now applying filter and removing "discarded" cells

```{r}
scae <- scae[, !scae$discard]
```

## Feature filtering

Removing all features that have a total count less than 5; with the exception of all alleles - all alleles are retained

```{r}
'%nin%' <- Negate('%in%')

filter_features <- function(scae){
  alleles <- grep("^ENS", rownames(scae), invert=TRUE, value=TRUE)
  other_genes <- rownames(scae) %nin% alleles
  features_keep <- rowSums(counts(scae)[other_genes, ]) > 5
  features_keep_names <- names(features_keep)[which(features_keep)]
  scae <- scae[c(features_keep_names, alleles), ]
  return(scae)
}

scae <- filter_features(scae)
dim(scae)
```

# Processing workflow

## Normalization

The data is normalized upon creating the *SingleCellAlleleExperiment* object

## Variance modeling

```{r, warning=FALSE}
gene_var <- modelGeneVar(scae)
plot(gene_var$mean, gene_var$total, pch=16, cex=0.5,
     xlab="mean of log-expression", ylab="variance of log-expression")
curfit <- metadata(gene_var)
points(curfit$mean, curfit$var, col="red", pch=16)
curve(curfit$trend(x), col='dodgerblue', add=TRUE, lwd=2) 
```

## Distribution of allele features

```{r, fig.asp=1}
p <- do.call(plot_grid, c(lapply(immune_features, function(feature){
  ggplot(as.data.frame(logcounts(scae)[feature, ]), aes(x=logcounts(scae)[feature, ])) +
    geom_histogram(color="black",
                   fill = "black",
                   breaks=seq(0, 8, by=0.10)) +
    labs(x = feature) +
    theme(axis.title.x = element_text(size = 8),
          axis.title.y = element_text(size = 8),
          plot.title = element_text(size = 8))}), ncol = 4))
print(p)
```

```{r}
immune_features <- rownames(scae[rowData(scae)$NI_I == "I", ])

plotExpression(scae, immune_features) +
  theme_bw() +
  theme(axis.text = element_text(size = 5),
        axis.title = element_text(size = 8, face = "bold"), 
        axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r, eval=FALSE}
saveRDS(scae, "../intermediate_files/scae.Rds")
```

# Session

```{r sessionInfo, cache = 0}
sessionInfo()
```
